---
title: "get_sens and plt_sens: Sensitivity Analysis for Unmeasured Confounding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{get_sens and plt_sens: Sensitivity Analysis for Unmeasured Confounding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5.5, dpi = 120,
  out.width = "100%", fig.align = "center",
  message = FALSE, warning = FALSE
)

# Every backend is a Suggests dependency, and two of them are GitHub-only, so
# each section is skipped rather than failed when its packages are missing.
has_lm   <- requireNamespace("sensemakr", quietly = TRUE)
has_cox  <- requireNamespace("survival", quietly = TRUE) &&
            requireNamespace("tipr", quietly = TRUE)
has_dml  <- requireNamespace("dml.sensemakr", quietly = TRUE)
has_iv   <- requireNamespace("iv.sensemakr", quietly = TRUE)
has_wrap <- requireNamespace("ggplotify", quietly = TRUE)
```

Every effect estimated from observational data rests on an assumption no data
can check: that nothing outside the model confounds treatment and outcome.
Sensitivity analysis replaces the yes/no question "is that true?" with a
quantitative one: **how strong would an unmeasured confounder have to be before
the conclusion changes?**

`get_sens()` answers that question through one of four established backends and
returns all four in the same shape. `plt_sens()` draws the matching figure.

```{r load}
library(causalR)
```

## Choosing a backend

| Your estimate | `method` | Backend | What it reports |
|---|---|---|---|
| Linear outcome (OLS) | `"lm"` | sensemakr | Partial \(R^2\), robustness value, bias-adjusted bounds |
| Cox model / hazard ratio | `"cox"` | tipr | Tipping point, E-value |
| Double machine learning | `"dml"` | dml.sensemakr | Robustness value, confounding bounds on the ATE |
| Instrumental variable | `"iv"` | iv.sensemakr | Robustness value, Anderson–Rubin bounds |

The two `sensemakr`-family backends are not on CRAN:

```r
remotes::install_github(c("carloscinelli/dml.sensemakr",
                          "carloscinelli/iv.sensemakr"))
```

## One output shape

Whichever backend runs, the result is a `sens_res` with the same four elements:

- **`$stats`** — one row per estimand, always the same 18 columns.
- **`$bounds`** — the confounding scenarios, always the same 12 columns
  (`NULL` when none was requested, and always `NULL` for `"cox"`).
- **`$fit`** — the fitted model (`lm`, `coxph`, `dml` or `iv_fit`).
- **`$sens`** — the backend's own sensitivity object.

Columns a backend does not produce are `NA` rather than absent, so results from
different backends stack with `rbind()`. Printing the object also echoes a
ready-to-paste `plt_sens()` call.

## Linear outcome: `method = "lm"`

```{r skip-lm, eval = !has_lm, echo = FALSE, results = "asis"}
cat("> **Not run.** This section needs sensemakr, not available here, so the ",
    "code below was skipped and produced no output or figure.\n", sep = "")
```

The Darfur data ask whether being directly harmed in an attack changes later
support for peace. The worry is that men were both more exposed and more
hawkish, so `female` is the natural yardstick for an unmeasured confounder.

```{r lm-fit, eval = has_lm}
adj <- c("age", "farmer_dar", "herder_dar", "pastvoted",
         "hhsize_darfur", "female", "village")

res_lm <- get_sens(sensemakr::darfur,
                   treat = "directlyharmed", outcome = "peacefactor",
                   adj_var = adj, bench_var = "female", method = "lm",
                   bench_args = list(k_treat = 1:3))
res_lm
```

Two numbers carry the argument. `rv_q` is the **robustness value**: the share of
residual variation in *both* treatment and outcome that a confounder would have
to explain to drive the estimate to zero. `rv_qa` is the same thing for the
confidence interval rather than the point estimate. `r2yd_x` is how much the
treatment itself explains, which puts the robustness value in context.

The `$bounds` rows answer the calibrated version: a confounder as strong as
`female`, twice as strong, three times as strong.

```{r lm-contour, eval = has_lm}
plt_sens(res_lm, type = "contour")
```

Each contour is a level of the adjusted estimate; the red dashed line is where
it reaches the threshold (zero by default). The triangle at the origin is the
unadjusted estimate and the diamonds are the benchmark scenarios. Since all
three benchmarks sit left of the critical line, a confounder up to three times
as strong as `female` would not overturn the result.

The extreme-scenario plot asks the same question from the other direction: if a
confounder explained essentially all of the residual outcome variation, how much
of the treatment would it need to explain?

```{r lm-extreme, eval = has_lm && has_wrap}
plt_sens(res_lm, type = "extreme", extreme_r2 = c(1, 0.75, 0.5))
```

## Cox model: `method = "cox"`

```{r skip-cox, eval = !has_cox, echo = FALSE, results = "asis"}
cat("> **Not run.** This section needs `survival` and `tipr`, not available here, so the ",
    "code below was skipped and produced no output or figure.\n", sep = "")
```

For a hazard ratio there is no partial \(R^2\) to report, so this backend gives
the two quantities that clinical journals ask for: the **tipping point** (how
strong the confounder–outcome association must be before the interval crosses
the null) and the **E-value** (how strong the confounder must be with *both*
exposure and outcome to explain the effect away).

```{r cox-fit, eval = has_cox}
lung <- stats::na.omit(survival::lung[, c("time", "status", "sex", "age", "ph.ecog")])
lung$status <- lung$status - 1L
lung$sex <- factor(lung$sex, labels = c("male", "female"))

res_cox <- get_sens(lung, treat = "sex", outcome = "status", time = "time",
                    adj_var = c("age", "ph.ecog"), method = "cox")
res_cox$stats[, c("estimate", "conf.low", "conf.high",
                  "evalue_point", "evalue_ci", "tip_effect")]
```

Note that `estimate` and the confidence limits are hazard ratios while
`std.error` is on the log-hazard scale — the same convention `summary.coxph()`
uses.

```{r cox-tip, eval = has_cox, fig.height = 5}
plt_sens(res_cox, type = "tip")
```

The dotted vertical line is the tipping point. The upper curve is the confidence
limit nearest the null, and it crosses 1 exactly there.

```{r cox-evalue, eval = has_cox, fig.height = 5}
plt_sens(res_cox, type = "evalue")
```

Each curve traces the confounder–exposure and confounder–outcome risk ratios
that would exactly explain the effect away. The marked points are where a curve
meets the diagonal — the E-value is the smallest pair of *equal* associations
that would suffice.

### Choosing `rare`

A hazard ratio only approximates a risk ratio when the outcome is rare.
`evalue_args$rare` decides how that is handled, and the same switch drives
`tipr`'s `hr_correction`:

```{r cox-rare, eval = has_cox}
rbind(
  common = get_sens(lung, treat = "sex", outcome = "status", time = "time",
                    adj_var = c("age", "ph.ecog"), method = "cox",
                    evalue_args = list(rare = FALSE))$stats,
  rare   = get_sens(lung, treat = "sex", outcome = "status", time = "time",
                    adj_var = c("age", "ph.ecog"), method = "cox",
                    evalue_args = list(rare = TRUE))$stats
)[, c("evalue_point", "evalue_ci")]
```

The default `rare = FALSE` applies the VanderWeele–Ding transform
\(RR = (1 - 0.5^{\sqrt{HR}}) / (1 - 0.5^{\sqrt{1/HR}})\), which shrinks the
ratio towards the null and so reports the smaller, more conservative E-value.
Set `rare = TRUE` only when the outcome really is rare over follow-up. This
differs from `tipr`'s own default, which leaves the hazard ratio untransformed.

A binary confounder is parameterised by its prevalence in each arm instead of by
a standardised mean difference:

```{r cox-binary, eval = has_cox}
get_sens(lung, treat = "sex", outcome = "status", time = "time",
         adj_var = c("age", "ph.ecog"), method = "cox",
         evalue_args = list(confounder = "binary",
                            exposed_prev = 0.5, unexposed_prev = 0.2)
)$stats[, c("tip_effect", "tip_n")]
```

## Double machine learning: `method = "dml"`

```{r skip-dml, eval = !has_dml, echo = FALSE, results = "asis"}
cat("> **Not run.** This section needs the GitHub-only `dml.sensemakr`, not available here, so the ",
    "code below was skipped and produced no output or figure.\n", sep = "")
```

With flexible learners in place of a linear adjustment, the sensitivity question
becomes how much residual variation a confounder would have to explain on each
side. `dml_args` controls the cross-fitting; the settings below are deliberately
small so this vignette builds quickly.

```{r dml-fit, eval = has_dml}
data("pension", package = "dml.sensemakr")
set.seed(20260921)
pen <- pension[sample(nrow(pension), 900L), ]

res_dml <- get_sens(pen, treat = "e401", outcome = "net_tfa",
                    adj_var = c("age", "inc", "educ", "fsize",
                                "marr", "twoearn", "pira", "hown"),
                    bench_var = "inc", method = "dml",
                    bench_args = list(k_treat = 1:2, bound = c(0.04, 0.03)),
                    dml_args = list(cf_folds = 2L, cf_reps = 1L,
                                    cf_seed = 20260921L, dirty_tuning = FALSE))
res_dml$stats[, c("estimand", "estimate", "conf.low", "conf.high",
                  "rv_q", "rv_qa")]
res_dml$bounds
```

DML bounds are *intervals*, not point adjustments: `adj_low`/`adj_high` bracket
the estimate and `adj_conf.low`/`adj_conf.high` bracket the confidence interval.
That is why `adj_estimate`, `adj_se` and `adj_statistic` are `NA` here — there
is no single adjusted number to report.

```{r dml-contour, eval = has_dml && has_wrap}
plt_sens(res_dml, type = "contour")
```

Raise `dml_args$cf_folds` and leave `dirty_tuning = TRUE` for real work; set
`cf_seed` so the cross-fitting is reproducible.

## Instrumental variables: `method = "iv"`

```{r skip-iv, eval = !has_iv, echo = FALSE, results = "asis"}
cat("> **Not run.** This section needs the GitHub-only `iv.sensemakr`, not available here, so the ",
    "code below was skipped and produced no output or figure.\n", sep = "")
```

Here the assumption at risk is exclusion: the instrument must affect the outcome
only through the treatment. `iv.sensemakr` asks how strongly a confounder would
have to be associated with the instrument before the Anderson–Rubin interval
stops excluding the null.

```{r iv-fit, eval = has_iv}
data("card", package = "iv.sensemakr")

res_iv <- get_sens(card, treat = "educ", outcome = "lwage",
                   instrument = "nearc4",
                   adj_var = c("exper", "expersq", "black", "south", "smsa",
                               "reg661", "reg662", "reg663", "reg664",
                               "reg665", "reg666", "reg667", "reg668",
                               "smsa66"),
                   bench_var = "black", method = "iv",
                   bench_args = list(k_treat = 1:3))
res_iv$stats[, c("estimand", "estimate", "conf.low", "conf.high",
                 "rv_qa", "xrv_qa")]
```

Three estimands are reported by default: the IV estimate itself, the first stage
(`fs`) and the reduced form (`rf`). `iv_args$parm` narrows that. `conf.low` and
`conf.high` are Anderson–Rubin limits, which is why there is no `std.error` —
the interval is not symmetric around the estimate and can be unbounded.

```{r iv-contour, eval = has_iv && has_wrap}
plt_sens(res_iv, type = "contour")
```

## Reading the numbers

The two headline quantities are on different scales and **must not be
compared**:

| | Scale | Range | Reads as |
|---|---|---|---|
| `rv_q`, `rv_qa`, `xrv_qa` | Partial \(R^2\) | \([0, 1)\) | Share of residual variation the confounder must explain |
| `evalue_point`, `evalue_ci` | Risk-ratio multiplier | \([1, \infty)\) | How strongly the confounder must be associated with both exposure and outcome |

There is no monotone mapping between them — the E-value depends on the effect
scale and on whether the outcome is rare — so they live in separate columns and
each is `NA` outside its own backend. A robustness value of 0.14 and an E-value
of 2.29 say nothing about each other.

## Benchmarking and manual scenarios

`bench_args` is shared by the `"lm"`, `"dml"` and `"iv"` backends. `k_treat`
(and `k_out`, which defaults to it) sets the multiples of `bench_var`; `bound`
adds one hand-specified scenario as `c(treatment-side R2, outcome-side R2)`,
mapped to whichever parameter names the backend uses.

```{r bench, eval = has_lm}
get_sens(sensemakr::darfur, treat = "directlyharmed", outcome = "peacefactor",
         adj_var = adj, bench_var = c("female", "pastvoted"), method = "lm",
         bench_args = list(k_treat = c(1, 2), bound = c(0.05, 0.10),
                           bound_label = "Worst case")
)$bounds[, c("bound_label", "r2_treat", "r2_out", "adj_estimate")]
```

`bench_var` must be a subset of `adj_var` — a yardstick you did not adjust for
is not a yardstick.

## Saving

`plt_sens()` follows the shared save contract: `save` is the last argument, both
`NULL` and `list()` skip saving, and a non-empty list is forwarded to
`RegR::save_plt()`. Width and height left out default to the size the figure was
pinned to.

```r
plt_sens(res_cox, type = "tip",
         save = list(filename = "Figure 3. Tipping point.pdf"))
```

## Things that will bite you

**Arguments that do not apply are an error, not a warning.** Passing
`dml_args` to `method = "lm"`, or `time` to anything but `"cox"`, stops with a
message naming the methods that do use it. Silently ignoring a backend list is
the easiest way to believe a setting took effect when it never did.

```{r unused, eval = has_lm, error = TRUE}
get_sens(sensemakr::darfur, treat = "directlyharmed", outcome = "peacefactor",
         adj_var = adj, method = "lm", dml_args = list(cf_folds = 10L))
```

**`EValue` is deliberately not used.** Its `estimate` methods clash with `lava`,
which `dml.sensemakr` pulls in through `caret`: once both namespaces are loaded,
`EValue::evalues.HR()` fails in either load order. The E-values here come from
`tipr::e_value()` plus the VanderWeele–Ding transform and reproduce
`EValue::evalues.HR()` to six decimal places on both settings of `rare`.

**Not every figure is a real ggplot.** The `"lm"` contour and both Cox figures
are drawn natively, so they accept further layers, scales and themes. The
`"dml"` and `"iv"` contours and the `"extreme"` plot are base-graphics figures
captured with `ggplotify::as.ggplot()`: they save and compose, but their
contents are a fixed grob, and because `gridGraphics` cannot emulate labels on
contour lines those two contours carry no level labels. The IV contour also
ignores `lim`, `threshold` and `contour_args`, since
`iv.sensemakr::ovb_contour_plot()` accepts none of them.

**dplyr verbs drop the `method` attribute** carried by `$stats` and `$bounds`.
Subset with `[` if you need it preserved.

## See also

`?get_sens`, `?plt_sens`, and the upstream references: Cinelli & Hazlett (2020)
for the partial \(R^2\) framework, VanderWeele & Ding (2017) for the E-value,
Chernozhukov et al. (2024) for DML sensitivity, and Cinelli & Hazlett (2022) for
the IV case.
